\plaintitle

### 2020-03-21 wip-design/paper.tex
Evaluation

\documentclass{sigchi}
\usepackage{xspace}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{xcolor}
\usepackage{tikz}
\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
    \draw (0,0) circle (1);
    \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
    \end{tikzpicture}}

@@

\subsection{Data sources}

@@

business team of a non-profit research organization, including a former journalist

\subsection{Research questions}

\paragraph{RQ1: Can end users explore data with The Gamma?}

\paragraph{RQ2: Can knowledge transfer between data sources?}

\paragraph{RQ3: Can users learn from just code samples?}

\subsection{Study design}
this is between-user user study

\subsection{Results}
it is good

\paragraph{Can end users explore data with The Gamma?}
Yes, the table shows the results.
People who spent between 10 and 25 minutes (average 18 minutes) working with The Gamma
Out of 13 people, all completed some work, 1 person completed partly and 6 were able
  to complete the task with small assistance (related to methods and syntax, as discussed below)

@@

we should use it as a training resource for students - would students get this?

\begin{quote}
\emph{``I don't think they'd need more than 5 minute video. This is a data
  source, this is what's in there.''}
\end{quote}

\begin{quote}
\emph{``for somebody who does not do coding or programming, this does not feel that daunting. it's not like you're giving me large screen full of code, which is reassurring" (~36)''}
\end{quote}

\paragraph{RQ2: Can knowledge transfer between data sources?}
expenditure task (#1-#5) and lords task (#9-#11) - is transfer from worldbank to expenditure and
from graph data to lords; the former transfers more directly, the latter less so (mainly because
of then and methods, which we discuss below)

\begin{quote}
  \emph{``I found it quite easy to translate what you showed us in the demo to the new dataset, i though it was quite easy to just get how that works''}
\end{quote}

\paragraph{RQ3: Can users learn from just code samples?}
in the worldbank task (#6-#9), users were given just a code sample and demo using no real
data source. They were able to complete based on just following and modifying code sample (to find other data).
(no problems they had were related to data source)

- "a video would just be this [code samples] anyway" (~19)
- "keystroke by keystroke video would be good"

\begin{quote}
  \emph{``I think providing one video of one example is good and maybe a couple of little examples of code wher people can see the kind of things you can do" (28:02)''}
\end{quote}

"more than one example of code would be useful because I'd be less confused about options"

\subsection{Further observations}

\paragraph{Making complex things possible hurts}
#11, #12, #13 all struggled with methods (they show in the completion lists)

@@

@@

\begin{table}
  \centering
  \begin{tabular}{l l l c l}
    \toprule
      & {\small \textit{Task}}
      & {\small \textit{Kind}} & {\small \textit{Completed}}
      & {\small \textit{Notes}} \\
    \midrule
    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid methods}\\
    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid methods}\\
    \bottomrule
  \end{tabular}
  \caption{Table captions should be placed below the table. We
    recommend table lines be 1 point, 25\% black. Minimize use of
    table grid lines.}~\label{tab:table1}
\end{table}

@@

>>>>

\documentclass{sigchi}
\usepackage{xspace}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{xcolor}
\usepackage{tikz}
\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
    \draw (0,0) circle (1);
    \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
    \end{tikzpicture}}
\newcommand*{\priorityc}[1]{
\resizebox{0.8em}{0.8em}{
  {\protect\tikz{ \protect
    \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
    } }}}

@@

\subsection{Data cube}

\subsection{Tabular data}

\subsection{Graph database}

@@

Our goal is to develop an easy-to-learn tool that journalists and other non-programmers can use
for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
we conduct a user study. We recruit participants among the operations and business team of a
non-profit research organization and investigate three research questions.

\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
We give participants one of four simeple data exploration tasks and see if they can complete the
task using The Gamma and, possibly, how much assistance they need.

\paragraph{RQ2: Can knowledge transfer between data sources?}
Our first hypothesis is that users familiar with the iterative prompting principle will be able to
use an unfamiliar data source. In two of the tasks, participants are shown one data source and
then asked to work with a different one.

\paragraph{RQ3: Can users learn from just code samples?}
Our second hypothesis is that users can learn through precolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants are given a code sample, but only a
minimal explanation of The Gamma environment.

\subsection{Study design}
We perform between-subjects study to evaluate the first experience of using The Gamma.
We recruit 13 participants (5 male, 8 female) from a business team
of a research lab working in various non-technical roles (project management,
partnerships, communications) including one former journalist.

We split participants into 4 groups. We first gave a brief 5 minute demonstration of The
Gamma and then asked participants to complete a task. Finally, we conducted a short semi-structured
group interview and later sent participants a follow-up questionnaire. The four tasks were:

\begin{itemize}
\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{expenditure} data source to compare the UK government spending
  on ``Public order and safety'' and ``Defence'' in terms of GDP.
\item \emph{Lodrs.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{lords} data source (a table with House of Lords
  members expenses) to find a members representing London with the highest travel costs.
\item \emph{Worldbank.} Participants were given a minimal explanation of The Gamma environment and
  a code sample using \emph{worldbank}. They were asked to solve a different task using
  the \emph{worldbank} data source.
\item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
  They were asked to solve a more complex problem using the same data source.
\end{itemize}

We let participants work independently, but offered guidance if they got stuck for longer period of
time. Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
\emph{worldbank} aims to answer RQ1 and RQ3. In \emph{olympics}, we test RQ1 using a more complex
data source and we ask further questions to explore their understanding of the data source.

\begin{table}
  \centering
  \begin{tabular}{l l l c l}
    \toprule
      & {\small \textit{Task}}
      & {\small \textit{Kind}} & {\small \textit{Done}}
      & {\small \textit{Notes}} \\
    \midrule
    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
    \bottomrule
  \end{tabular}
  \caption{Overview of the work completed by individual participants. The marks denote:
    \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
  \label{tab:tasks}
  \vspace{-1em}
\end{table}

\subsection{Results}
Table~\ref{tab:tasks} summarizes the work done by the study participants.
For each participant, we record the task, the kind of data source used in the task and the
level of completion. For participants who needed assistance, the notes section details the help
given. We analyse the hints in the next section. Table~\ref{tab:quest} shows the results of the
follow-up questionnaire.

\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
Three results allow us to answer RQ1 in the affirmative. First, 7 out of 9
participants agree or strongly agree that they ``found the system easy to use''. Second,
participants spent between 10 and 25 minutes (average 17 minutes) working with The Gamma and
all 12 out of 13 participants completed the task; 6 participants required some assistance,
but 3 of those faced one issue (discussed later) that could be covered in the introductory
presentation. Third, a number of participants shared positive comments in the semi-structured
interview.

Participant \#3 found the system simple, but also points out an issue about data provenance,
which we discuss later:

@@

Similarly, participant \#2, notes that The Gamma alleviated their unease about code:

\begin{quote}
\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
  It's not like you're giving me large screen full of code, which is reassurring.''}
\end{quote}

Finally, participant \#5 suggested the system could be used as an educational tool for teaching
critical thinking with data. They answer a follow-up question about what training materials would
the students need as follows:

\begin{quote}
\emph{``I don't think they'd need more than 5 minute video (..) this is the data
  source, this is what's in there.''}
\end{quote}

\paragraph{RQ2: Can knowledge transfer between data sources?}
Our study does not conclusively answer RQ2. There is some evidence in favor of a positive answer.
In the practical part, two of the tasks (\emph{expenditure} and \emph{lords}) used a different data
source in the introductory presentation than the one that the participants were asked to use.
Participants were able to complete those tasks, although \emph{lords} has been more challenging
as it involves a more complex data source. In the interview, participant \#2 also gives a positive
answer:

\begin{quote}
\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.
   I though it was quite easy to just get how that works.''}
\end{quote}

Negative evidence is offered by the follow-up questionnaire. When asked
whether they would know how to approach a task using a new data source, 5 out of 9 participants
diagree or strongly disagree that they would know how to approach it without any further guidance.
Participants did not believe that knowledge can easily transfer to another
data source.

\paragraph{RQ3: Can users learn from just code samples?}
A positive answer to RQ3 is supported by the \emph{worldbank} task results, the follow-up questionnaire
and interview comments. In the task, participants were given only a
minimal demo of the iterative prompting principle together with print-out of 2 code samples.
All three were able to complete a related task using the same data source. In the follow-up
questionnaire, only 2 out of 9 participants disagree or strongly disagree that
they would know how to approach a task using an unfamiliar data source when given ``a number
of code samples''.

In the semi-structured interview, participants were asked what would be the most useful format
of educational material about The Gamma (code samples, video tutorials, etc.).
Participant \#7 noted that \emph{``a video would just be this [i.e.~a code sample] anyway''}, while
participant \#13 (former journalist) answered:

\begin{quote}
  \emph{``I think providing one video of one example is good and maybe a couple of little examples
  of code wher people can see the kind of things you can do.''}
\end{quote}

This is aligned with our design goal. Once the user understands the iterative prompting principle
(which can be done through a video tutorial), they can learn how to use any specific data source
just from code samples.

\subsection{Further observations}
In this section, we briefly discuss a number of observations about The Gamma design that
emerged from the study, some of which suggest ways of improving the system.

\paragraph{Making complex things possible may hurt}
As discussed earlier, The Gamma language supports operations such as \ikvd{take(5)}. Most type
providers never generate those, but the provider for working with tabular data is an exception.
When filtering data, the provider allows specifying a condition on numerical attributes such as
\ikvd{olympics.\textquotesingle filter data\textquotesingle.\textquotesingle Year is greater than\textquotesingle(2004)}.

Three participants (\#11, \#12, \#13) struggled to complete a task, because they
initially attempted to use those operations. However, those violate the iterative prompting
principle as one cannot type `.' after \ikvd{\textquotesingle Year is greater than\textquotesingle}.
This suggests that we should avoid operations in type providers for data access, even if it
limits the expressive power of the type provider.

@@

\begin{table}
  \centering
  \begin{tabular}{p{19em} c c}
    \toprule
      {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
    \midrule
    \small Considering your experience \& possible use of The Gamma:\\
    \small \quad I found the system easy to use. & \small 2.11 & \small 0.87\\
    \small \quad Journalists will be able to use it to analyse data. & \small 2.67 & \small 1.05\\
    \small \quad Readers will be able to critically examine the analyses. & \small 2.44 & \small 0.68\\
    \small Thinking about the usability of the system:\\
    \small \quad The resulting programs are easy to understand. & \small 2.33 & \small 0.82\\
    \small \quad The text editor for creating them was easy to use. & \small 2.33 & \small 0.94\\
    \small \quad Finding the right item in the auto-complete was easy. & \small 2.56 & \small 1.42\\
    \small I would know how to approach a task using a new data source\\
    \small \quad If given a comprehensive video tutorial. & \small 2.00 & \small 1.24\\
    \small \quad If given a number of code samples. & \small 2.78 & \small 1.13\\
    \small \quad Without any further guidance. & \small 3.56 & \small 1.17\\
    \bottomrule
  \end{tabular}
  \caption{Summary of follow-up questionnaire responses using
    a 5-point Likert scale (1 = strongly agree, 5 = strongly disagree) over 9 subjects.}
  \label{tab:quest}
\end{table}

@@

@@

Data provenance (where data comes from)
What are the data sources availbale (how to start)?
